Global patterns and drivers of vertebrate verticality

### 2025-01-22 verticality_future_mkd.Rmd
update manuscript

@@

In forest ecosystems, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy, i.e., vertical stratification [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition across vertical strata akin to multi-floored apartments [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. This vertical stacking of species has become an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. As such, vertical stratification may be a mechanistic link to positive relationships between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023]. However, canopy height does not consistently impact species richness at regional to global scales, with notable declines in the strength of the relationship at poleward latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019].

Tall forests contain steep vertical climate gradients characterized by lower thermal variability and greater moisture toward the ground [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. At temperate and boreal latitudes, low temperature and moisture in the canopy may exceed physiological thresholds and constrain species to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021]. In wet tropical forests, physiological constraints decline, which may allow species to stratify across the entire vertical habitat gradient [@oliveiraVerticalStratificationInfluences2019]. We therefore expect a globally varying relationship between canopy height and verticality pending baseline regional conditions from positive in tropical to neutral in temperate and boreal systems.

Spatial variation may also occur in the relationship between verticality and species richness [@oliveiraVerticalStratificationInfluences2019]. Though prior studies have shown positive relationships between verticality and richness for select taxa, the strength of this relationship is variable and generally more positive in tropical regions where climatic conditions pose fewer physiological constraints on arboreal strategies [@cabralFunctionalTraitsPhylogeny2022; @oliveiraVerticalStratificationInfluences2019]. Niche expansion, characterized by vertical partitioning of the foerst, may allow this pattern to occur. In contrast, the relationship between verticality and richness may be weak where taller canopies do not facilitate higher verticality.

As the climate changes, constraints on arboreal life strategies may shift with potential impacts on assemblage verticality and its association with species richness. In tropical lowlands, increasing temperatures and declining dry season precipitation (*cite*) may pose physiological challenges for arboreal animals and cause them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. These scenarios would reduce assemblage verticality in the lowlands, akin to studies that show shifts in ecological processes such as climate-induced shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. At poleward latitudes, increasing minimum temperatures could reduce climatic constraints on arboreal species and facilitate range shifts that increase verticality.

@@

Despite decatdes of ecological inquiry into the relationship between forest height and species richness, little is known as to whether tall forests ubiquitously support niche filling and thus promote greater species coexistence per unit of area (high alpha or gamma diversity) To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of vertebrates (`r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians) with climate and vegetation data to examine global patterns and drivers of verticality in wooded biomes. We then examine whether the dynamic of filling vertical niche space and its effects of species richness is informed by the relationship between verticality, canopy height, and climate. Finally, we predict how climate change may impact global patterns of verticality.

@@

We mapped verticality and richness of vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] using 50 km grid cells and a cylindrical equal area projection (Supplementary methods). Grid cells with fewer than five species were excluded. Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022]. For birds, we used breeding plus resident ranges. To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024]. For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014]. We classified foraging below the water surface, around the water surface, and on the ground as 0.5, in the understory as 0.667, mid-high canopy as 0.8337, and in or above the upper canopy as 1. If species used multiple vertical habitats, vertical scores were averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that did not occur in terrestrial habitats were excluded from the analysis.

To examine spatial patterns of verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality ($SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool). We defined species pools as species within all wooded habitats in each biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality given random chance for each grid cell using 100 simulations.

@@

We obtained present and future global climate data from CHELSA v2.1 for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density). Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality [@oliveiraVerticalStratificationInfluences2019]. All environmental data were resampled to the resolution and projection of the species data.

@@

### Variable selection
We examined relationships between mean or SES verticality and the ten environmental variables: canopy height, vegetation density, vegetation complexity, maximum temperature of the warmest month, minimum temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, precipitation seasonality, and climate velocity.  To linearize the relationships between predictor and response variables, we log-transformed climate velocity and precipitation of the driest month. We then scaled all predictor variables to mean zero and variance one. To reduce impacts of collinearity on the model, we examined VIF factors of all variables and sequentially removed one variable at a time until all VIFs were less than 5 [@wartonEcostatsDataAnalysis2022]. We chose variables to remove based on a combination of which had the highest VIF and which were most likely to directly impact physiological limits to arboreality. For example, we excluded precipitation seasonality because it was strongly correlated with dry season precipitation and has a less direct impact on physiological limits to verticality than dry season precipitation. This procedure resulted in a final set of seven variables that included canopy height, vegetation density, maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the wettest month, precipitation of the driest month, and climate velocity. The additive effects of these variables were included as fixed effects in the model in addition to a quadratic term for maximum temperature of the warmest month, as we expected that verticality may increase with temperature up to a certain point but then decline under extremely high temperatures.

@@

We used spatial generalized linear mixed-effects models (GLMMs) to evaluate the impact of environmental variables on mean and SES verticality. Spatial models were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which fits models using maximum marginal likelihood estimation through template model builder (TMB) and uses the stochastic partial differential equation (SPDE) approach to approximate spatial Gaussian random fields. [@lindgrenExplicitLinkGaussian2011] as implemented in R-INLA [@lindgrenBayesianSpatialModelling2015]. sdmTMB is also able to fit models that include a spatially varying coefficient (SVC), which allow the coefficient of a predictor variable to vary across the study extent, and are better able to detect complex species-environment relationships [@doserGuidelinesUseSpatially2024; @andersonSdmTMBPackageFast2024].

@@

where $y_s$ is mean or SES verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected mean or SES verticality, and $𝑓^{-1}$ represents an inverse link function. We modeled SES verticality as a normal distribution and mean verticality as a beta distribution with a logit link function because values are constrained between 0 and 1. $𝜲^{main}_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients. $α_g$ represents a random intercept defined by biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017], which accounts for potential variation in verticality drive by global evolutionary patterns. $⍵_s$ represents a spatial random field at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates or random intercept. $𝜲^{svc}_sζ_s$ represents a spatially varying coefficient, where $𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$). Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance. Gaussian random fields were approximated for models of mean and SES verticality for each taxa using a spatial mesh optimized such that the edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019] (Supplementary methods). The model is specified such that both spatial terms share the same range parameter.

@@

For each taxonomic group and type of verticality (i.e., SES or mean verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structures [@zuurMixedEffectsModels2009]. The base model only included a spatial random field. Subsequent models additionally included biogeographic realm [sensu @dinersteinEcoregionBasedApproachProtecting2017] as a random intercept, a spatially varying coefficient for canopy height, and the combination of both. Including realm as a random intercept accounts for evolutionary differences among biogeographic realms that could lead to differences in verticality. The spatially varying coefficient allows the impact of canopy height on verticality to vary globally and allows us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality. We compared models using AIC and cross-validation. AIC is frequently used to identify the most parsimonious model [e.g., @barnettImprovingEstimatesSpecies2021]; however, the metric is not completely appropriate for hierarchical models (see Commander et al. 2022 `r word_comment(comment = "check citation", highlight = "Commander et al. 2022")`). Therefore, we additionally used cross-validation to evaluate out-of-sample predictive ability (*again check citation Commander et al. 2022*). We randomly assigned each data point to one of five folds, and then iteratively fit the model using four of the five folds while one fold was withheld for testing. The sum of the log-likelihoods for each left-out fold provides a measure of predictive ability that is compared across models. Cross validation was conducted using the ‘sdmTMB_cv’ function from the sdmTMB package. Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Tabe S\@ref(tab:mod-compare-tab)). Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model that included both realm and the SVC.

@@

## Relationship between verticality and richness

To examine whether the relationship between verticality and canopy height mediates the relationship between verticality and richness, we calculated Pearson correlations between mean or SES verticality and richness for each taxon considering all points and only points where canopy height had a positive significant effect on verticality.

@@

To examine potential impacts of climate change on assemblage verticality, we used the models to predict mean and SES verticality under a future climate scenario. We held vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density. We subtracted predictions to the present-day climate data used to train the model from future predictions and mapped the resulting differences. To standardize measures of change across taxa and between mean and SES verticality, we calculated the percent relative change between present and future predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell. Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.

@@

Verticality exhibited a latitudinal gradient across taxa characterized by high verticality in the tropics and lower verticality toward higher latitudes (Fig. \@ref(fig:vert-map-global)). This pattern was largely driven by declines in the proportion of arboreal species toward the poles. Patterns of SES verticality mirrored those of mean verticality and highlight declines in arboreal strategies outside tropical moist forests (Fig. S\@ref(fig:vert-biome)). Within tropical and subtropical biomes, verticality was highest in moist forests and lowest in dry forests (Fig. \@ref(fig:vert-biome)). In moist forests, verticality of bird and mammal assemblages was consistently high, with an average of `r birds.arb.tropicalmoist`% of birds and `r mammals.arb.tropicalmoist`% of mammals exhibiting arboreal strategies. In contrast, amphibian and reptile verticality exhibited clear hotspots, with reptile verticality peaking in Madagascar and Malaysian and Indonesian forests and amphibian verticality peaking in the Brazilian Atlantic forest and eastern Madagascar.

In temperate biomes, verticality was similar across forest types but declined in less dense wooded habitats (e.g., savannas and shrublands) (Fig. S \@ref(fig:vert-biome)). Variation in the proportion of species exhibiting fossorial strategies drove differences in patterns of assemblage verticality among vertebrate classes. Amphibians exhibited the highest proportion of fossorial strategies in temperate biomes (Fig. \@ref(fig:vert-map-global)L), though temperate rainforests (e.g., the Pacific-Northwest of North America) were a notable exception. Mammals exhibited a similar pattern, though the proportion of fossorial species increased to a lesser extent (Fig. \@ref(fig:vert-map-global)F). In contrast, the proportion of fossorial reptiles declined in temperate regions. Low reptile verticality was therefore driven by declines in the proportion of arboreal species and increases in the proportion of terrestrial species. Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal verticality gradient.

@@

When considering all points, Pearson correlations indicated strong positive relationships between SES or mean verticality and species richness for birds and mammals, a weak positive relationship for amphibians, and nearly no relationship for reptiles (Fig. \@ref(fig:corplot)). When we restricted the correlation to only include points where canopy height had a significant positive effect on verticality, the strength of the relationship between SES verticality and richness increased for amphibians and SES verticality of birds, but declined for mammals, reptiles, and mean verticality of birds.

@@

Our results indicate globally consistent macroscale patterns of verticality across taxa that are characterized by declining verticality toward higher latitudes and correspond with previous studies that are more limited in taxonomic or spatial extent [@bashamVerticalStratificationPatterns2023a; @oliveiraVerticalStratificationInfluences2019]. Though climate conditions influenced verticality, they were neither able to predict extremely high or low verticality nor spatial variation in the influence of canopy height on verticality. Furthermore, regions exhibiting positive relationships between canopy height and verticality did not consistently exhibit stronger positive relationships between canopy height and species richness. We propose that the rate of niche packing across evolutionary time spans rather than expansion across tall forests may explain both the limited capacity for climate alone to predict high levels of verticality and the lack of the three-way positive relationship between canopy height, verticality, and species richness for some taxa. Given the potentially large role of evolutionary patterns in global distributions of verticality, our models may underpredict the impacts of climate change on vertical assemblage structure.

>>>>

# Abstract
_Aim_: Vertical stratification of species from the ground to canopy is posited as a mechanistic link explaining positive relationships between canopy height and species richness. A key question is understanding how climate and vegetation impact verticality and spatial variation in verticality-diversity patterns. We examine these relationships and then project models of verticality under future climate conditions to examine how climate change may impact vertical community structure.

_Location_: Global

_Time period_: 1981-2010 to 2071-2100

_Major taxa studied_: Vertebrates

_Methods_: We modeled impacts of climate and vegetation on vertebrate verticality globally using spatial generalized linear mixed models with a spatially varying coefficient for canopy height. We projected models to future climate conditions to examine impacts of climate change on vertical community structure. We also examined global and biome-specific relationships between verticality and species richness.

_Results_: Verticality patterns are globally consistent across vertebrate taxa, declining from low to high latitudes. Low precipitation and temperature constrain verticality outside moist tropical regions. Globally, positive relationships occurred between canopy height and verticality and between verticality and richness (except for reptiles). Regionally, canopy height exhibited largely neutral impacts on verticality, while verticality-richness relationships varied from negative or weakly positive in temperate and boreal biomes to generally positive in tropical biomes. Model projections indicated relatively small changes in verticality under future climate conditions.

_Main conclusions_: Physiological constraints to arboreal lifestyles limit verticality in cold and dry macroclimate conditions. Globally, vertical stratification across increasingly tall forests (niche expansion) contributes to high richness in tropical regions. Regionally, climatic constraints limit positive verticality-richness relationships in higher latitudes, while niche packing largely contributes to positive verticality-richness relationships in tropical regions. Small changes to vertical community structure at a global scale could be caused by arboreal species increasing their use of understory habitat, becoming extirpated, or shifting geographic ranges to higher latitudes.

@@

In forest ecosystems, species assemblages stratify vertically along resource and climate gradients that extend from beneath the ground to the top of the canopy, a phenomenon called vertical stratification [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall, multi-layered forests increase both habitat volume and niche opportunities, enabling species to partition across vertical strata akin to multi-floored apartments (Fig. \@ref(fig:concept-fig)a;[@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]). This vertical stacking of species has been described as a mechanistic link between canopy height and species richness [@culbertInfluenceVerticalHorizontal2013; @fengForestCanopyHeight2020; @gillespieDistributionBirdSpecies2001; @oliveiraVerticalStratificationInfluences2019; @wagnerRemotelySensedTree2023], and is thought to contribute to the exceptionally high biodiversity in the tropics via niche expansion processes [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. Thus, as the amount of habitat space increases with forest height, species may be added to the assemblage, increasing richness relative to shorter forests (Fig. \@ref(fig:concept-fig); @pellissierNichePackingExpansion2018; @pigotFunctionalTraitsReveal2016). However, canopy height does not consistently impact species richness at regional to global scales, with weaker relationships observed at high latitudes where verticality is lower [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019]. Whether verticality plays a central role in driving richness gradients globally remains an open question, and surprisingly little progress has been made in assessing whether this relationship holds across multiple taxonomic groups with differing sensitivities to climate variability.

Tall forests exhibit steep vertical climate gradients, characterized by high maximum temperatures, low minimum temperatures, and reduced moisture availability in the canopy and lower thermal variability and higher moisture availability toward the ground [@scheffersTropicalMountainPasses2018; @leahyVerticalNicheElevation2021; @defrenneGlobalBufferingTemperatures2019]. In wet tropical forests, macroclimate conditions produce warmer and wetter canopy microclimates, which reduce thermal and hydric stress and may allow species to stratify across the entire vertical habitat gradient, partitioning niche space including nesting and food resources [@oliveiraVerticalStratificationInfluences2019]. Taller canopies in these regions could thus lead to higher verticality and richness (Fig. \@ref(fig:concept-fig)). In contrast, at temperate and boreal latitudes, macroclimate conditions, including low temperature and precipitation, may cause canopy microclimates to exceed physiological thresholds for many species, often constraining them to lower vertical positions [@oliveiraVerticalStratificationInfluences2019; @bakenSalamanderArborealityLimited2021]. As a result, taller canopies in these regions are not necessarily associated with higher verticality and richness as harsh macroclimate conditions limit the availability of vertical habitat space (Fig. \@ref(fig:concept-fig)). We therefore expect a globally varying relationship between canopy height and verticality as well as between verticality and species richness, pending baseline regional conditions from positive in tropical to neutral in temperate and boreal ecosystems.

Since verticality is influenced by both climate and vegetation structure, changing climatic conditions may alter constraints on arboreal life strategies, potentially reshaping assemblage verticality and its association with species richness. In tropical lowlands, novel high temperatures [@trewNovelTemperaturesAre2024] may impose physiological stress on arboreal animals, leading them to disperse toward higher elevations, lower vertical positions, or become locally extirpated [@scheffersIncreasingArborealityAltitude2013; @oliveiraVerticalStratificationInfluences2019]. These scenarios would reduce assemblage verticality in the lowlands, mirroring documented climate-driven shifts in assemblage thermal tolerance [@feeleyClimatedrivenChangesComposition2020; @lajeunesseTemporalAnalysisGBIF2023; @chustCrossbasinCrosstaxaPatterns2024; @borderieuxExtinctionDrivesRecent2024]. In contrast, at higher latitudes, warming minimum temperatures may alleviate climatic constraints on arboreal species, promoting range expansions that could increase assemblage verticality. These contrasting dynamics underscore the need for further investigation into how climate-driven changes influence the vertical structure of assemblages and its impacts on biodiversity gradients.

Despite decades of ecological inquiry into the relationships between verticality and species richness, global patterns of verticality, the environmental factors driving verticality, and the mechanistic link between verticality and species richness have not been comprehensively described for terrestrial vertebrates. To improve our understanding of the nuances of the relationships between climate, vegetation structure, verticality, and species richness, we integrate range maps of vertebrates (`r nbirds` birds, `r nmammals` mammals, `r nrept` reptiles, and `r namph` amphibians) with climate and vegetation data to examine global patterns and drivers of verticality in wooded biomes, as well as relationships between verticality and species richness. We ask 1) how do climate and vegetation structure impact verticality and does the impact of canopy height vary spatially, 2) how do relationships between verticality and richness vary across biomes, and 3) how might climate change impact global patterns of verticality?

@@

@@

We mapped global verticality and richness of vertebrate assemblages in wooded habitats [@dinersteinEcoregionBasedApproachProtecting2017] within 50-km grid cells and using a cylindrical equal area projection (Supplementary Methods). Grid cells with fewer than five species were excluded, because they may reduce the performance of null models. Range maps were obtained from the IUCN red list for mammals and amphibians [@iucnIUCNRedList2023], BirdLife International for birds [@birdlifeinternationalandhandbookofthebirdsoftheworldBirdSpeciesDistribution2022; using breeding plus resident ranges], and the Global Assessment of Reptile Distributions (GARD) for reptiles [@rollGlobalDistributionTetrapods2017; @caetanoAutomatedAssessmentReveals2022].

To calculate the verticality of each assemblage (i.e., grid cell), we assigned each species a verticality score between zero and one and took the average verticality of all species whose ranges overlapped with the grid cell. For mammals, reptiles, and amphibians, verticality scores were based on their use of fossorial (0), terrestrial (0.5), and arboreal (1) habitats as defined in the TetrapodTraits database [@mouraPhylogenyinformedCharacterisationGlobal2024]. For birds, we calculated verticality scores based on their use of different foraging strata as identified in the EltonTraits database [@wilmanEltonTraitsSpecieslevelForaging2014], which provides a finer scale vertical classification than the dataset from Moura et al. [-@mouraPhylogenyinformedCharacterisationGlobal2024]. We classified foraging below the water surface, around the water surface, and on the ground as 0.5, in the understory as 0.667, mid-high canopy as 0.8337, and in or above the upper canopy as 1. If species used multiple vertical habitats, vertical scores were averaged. For example, if a species occurred in terrestrial (0.5) and arboreal (1) habitats, it would receive a verticality score of 0.75. Species that occupied only aerial vertical niches or that did not occur in terrestrial habitats were excluded from the analysis.

To examine spatial patterns of assemblage verticality independently of species richness, we calculated the standard effect size (SES) of mean verticality: $SES = (x_i - \bar{x})/sd(x)$, where x is a vector of expected mean verticality scores given random selection of species from a species pool. We defined species pools as all species inhabiting wooded habitats within a given biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017] and generated a null distribution of expected verticality for each grid cell by conducting 100 simulations with random samples matching the observed richness. Highly positive SES verticality scores represent assemblages having much higher verticality than expected given the species richness, while low SES verticality scores represent assemblages having much lower verticality than expected given the species richness.

@@

We obtained present and future global climate data from CHELSA v2.1 (30 arc-second resolution) for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density). Additionally, we included historical climate velocity (0.25° resolution), which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and influences global patterns of verticality [@oliveiraVerticalStratificationInfluences2019]. All environmental data were resampled to the resolution and projection of the species data.

@@

### Variable selection
We examined relationships between mean or SES verticality and the ten environmental variables: canopy height, vegetation density, vegetation complexity, maximum temperature of the warmest month, minimum temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, precipitation seasonality, and climate velocity. To linearize the relationships between predictor and response variables, we log10-transformed climate velocity and precipitation of the driest month. We then scaled all predictor variables to mean zero and variance one. 
To reduce impacts of collinearity on the model, we examined VIF factors of all variables and sequentially removed one variable at a time until all VIFs were less than five [@wartonEcostatsDataAnalysis2022]. We chose variables to remove based on a combination of which had the highest VIF and which were most likely to directly impact physiological limits to arboreality. For example, we excluded precipitation seasonality because it was strongly correlated with dry season precipitation and has a less direct impact on physiological limits to verticality than dry season precipitation. This procedure resulted in a final set of seven variables that included canopy height, vegetation density, maximum temperature of the warmest month, minimum temperature of the coldest month, precipitation of the wettest month, precipitation of the driest month, and climate velocity. The additive effects of these variables were included as fixed effects in the model in addition to a quadratic term for maximum temperature of the warmest month, as we expected that verticality may increase with temperature up to a certain point but then decline under extremely high temperatures.

@@

We used spatial generalized linear mixed-effects models (GLMMs) to evaluate the impact of environmental variables on mean and SES verticality. Spatial models were fit using the R package sdmTMB [@andersonSdmTMBPackageFast2024], which fits models using maximum marginal likelihood estimation through template model builder (TMB) and uses the stochastic partial differential equation (SPDE) approach to approximate spatial Gaussian random fields [@lindgrenExplicitLinkGaussian2011] as implemented in R-INLA [@lindgrenBayesianSpatialModelling2015]. sdmTMB is also able to fit models that include a spatially varying coefficient (SVC), which allow the coefficient of a predictor variable to vary across the study extent and are better able to detect complex species-environment relationships [@doserGuidelinesUseSpatially2024; @andersonSdmTMBPackageFast2024].

@@

where $y_s$ is mean or SES verticality at location $s$ defined by x and y coordinates, $𝜇_s$ is the expected mean or SES verticality, and $𝑓^{-1}$ represents an inverse link function. We modeled SES verticality as a normal distribution and mean verticality as a beta distribution with a logit link function because values are constrained between 0 and 1. $𝜲^{main}_s$ is a vector of main effect covariates and 𝛽 represents a corresponding vector of estimated main effect coefficients. $α_g$ represents a random intercept defined by biogeographic realm [@dinersteinEcoregionBasedApproachProtecting2017], which accounts for potential variation in verticality driven by global evolutionary patterns. $⍵_s$ represents a spatial random field at location $s$, and describes additional spatial variation in the data not explained by the environmental covariates or random intercept. $𝜲^{svc}_sζ_s$ represents a spatially varying coefficient, where $𝜲^{svc}$ represents the vector of spatially varying parameter values and $ζ_s$ represents the spatially varying coefficients, which are described using a random field such that the coefficients follow a multivariate normal distribution with mean zero and covariance ($𝚺_ζ$).

Both the spatial random field and the spatially varying coefficients are modeled as Gaussian random fields with a covariance matrix modeled by the Matérn covariance function, which defines the decay rate of correlation with distance. Gaussian random fields were approximated for models of mean and SES verticality for each taxa using a spatial mesh optimized such that the edge length was as large as possible to reduce computational time while remaining less than five times the spatial range of the model [@andersonSdmTMBPackageFast2024; @krainskiAdvancedSpatialModeling2019] (Supplementary methods). The model is specified such that both spatial terms share the same range parameter.

@@

For each taxonomic group and type of verticality (i.e., SES or mean verticality), we used restricted maximum likelihood (REML) to fit models with four different random effects structure [@zuurMixedEffectsModels2009]. The base model only included a spatial random field. Subsequent models additionally included biogeographic realm (sensu @dinersteinEcoregionBasedApproachProtecting2017) as a random intercept, a spatially varying coefficient for canopy height, and the combination of both. The SVC allows the impact of canopy height on verticality to vary globally and allows us to test the hypothesis that canopy height has a positive effect on verticality only where climatic conditions support arboreality.

We compared models using AIC and cross-validation. AIC is frequently used to identify the most parsimonious model (e.g., @barnettImprovingEstimatesSpecies2021); however, the metric is not completely appropriate for hierarchical models [@commanderShadowModelHow2022]. Therefore, we additionally used cross-validation to evaluate out-of-sample predictive ability [@commanderShadowModelHow2022]. We randomly assigned each data point to one of five folds, and then iteratively fit the model using four of the five folds while one fold was withheld for testing. The sum of the log-likelihoods for each left-out fold provides a measure of predictive ability that is compared across models. Cross validation was conducted using the ‘sdmTMB_cv’ function from the sdmTMB package. Models that included both a spatially varying coefficient for canopy height and geographic realm as a random intercept consistently had the lowest AIC value across all models, while model performance varied based on cross validation (Tabel S\@ref(tab:mod-compare-tab)). Given consistent performance based on AIC and our interest in the SVC, all predictions were made based on the model that included both realm and the SVC.

@@

## Relationships between verticality and richness

We fit linear models of richness as a function of SES or mean verticality for all assemblages as well as for assemblages in each biome [@dinersteinEcoregionBasedApproachProtecting2017]. Biomes with fewer than 30 assemblages were not included.

@@

To examine potential impacts of climate change on assemblage verticality, we used the models to predict mean and SES verticality under a future climate scenario. We held vegetation variables constant at present-day values due to lack of global predictions of future canopy height and vegetation density. We subtracted predictions to the present-day climate data used to train the model from future predictions and mapped the resulting differences. To standardize measures of change across taxa and between mean and SES verticality, we calculated the relative percent change between present and future predictions for mean and SES verticality as $(verticality_{future,t,i} - verticality_{present,t,i})/(verticality_{max,t} - verticality_{min, t})*100$, where t indicates a given taxon and i indicates a given grid cell. Negative values indicate predicted declines in verticality and positive values indicate predicted increases in verticality.

@@

The standard effect size (SES) of verticality and mean verticality exhibited latitudinal gradients across taxa characterized by high verticality in the tropics and lower verticality toward poleward latitudes (Fig. \@ref(fig:vert-map-global)). These patterns were largely driven by declines in the proportion of arboreal species toward the poles (Fig. \@ref(fig:vert-map-global)).

Within tropical and subtropical biomes, verticality across taxa was highest in moist forests and lowest in dry forests (Fig. S\@ref(fig:vert-biome)). In moist forests, verticality of bird and mammal assemblages was consistently high globally, with an average of `r birds.arb.tropicalmoist`% of birds and `r mammals.arb.tropicalmoist`% of mammals exhibiting arboreal strategies. In contrast, amphibian and reptile verticality exhibited clear hotspots, with reptile verticality peaking in Madagascar and Indo-Malay forests and amphibian verticality peaking in the Brazilian Atlantic forest and eastern Madagascar (Fig. \@ref(fig:vert-map-global)).

In temperate biomes, verticality was lower in less-dense wooded habitats (e.g., savannas and shrublands) (Fig. S\@ref(fig:vert-biome)). Additionally, the proportion of species exhibiting fossorial strategies largely influenced differences in verticality among vertebrate classes in temperate regions. Amphibians exhibited the highest proportion of fossorial strategies in temperate biomes (Fig. \@ref(fig:vert-map-global)L), followed by mammals (Fig. \@ref(fig:vert-map-global)F), although amphibians in temperate rainforests (e.g., the Pacific-Northwest of North America) had relatively high verticality. In contrast, the proportion of fossorial reptiles declined in temperate regions, with low verticality driven by declines in the proportion of arboreal species and increases in the proportion of terrestrial species. Though no birds exhibit fossorial strategies, a reduction in the proportion of arboreal species still produced a steep latitudinal verticality gradient (Fig. \@ref(fig:vert-map-global)).

@@

Linear models indicated globally positive relationships between richness and mean or SES verticality across all taxa (Fig. \@ref(fig:richness-mods)). However, R^2^ values indicated generally weaker global relationships for reptiles and amphibians than for birds and mammals. When restricted to individual biomes, boreal and temperate regions generally exhibited negative or weak positive slopes, while tropical regions generally exhibited stronger positive slopes (Fig. \@ref(fig:richness-mods), S\@ref(fig:cor-biome)). However, reptiles deviated from this pattern, exhibiting weak positive or negative slopes between richness and verticality in the tropics (Fig. \@ref(fig:richness-mods), S\@ref(fig:cor-biome)).

@@

Understanding global patterns and drivers of verticality is key to understanding how verticality mediates geographic patterns of species richness. Our results reveal globally consistent macroscale patterns of verticality across taxa in wooded biomes, with epicenters of verticality in the wet tropics and declining verticality toward the poles (Fig. \@ref(fig:vert-map-global)).

## Climatic drivers of verticality

Climate influences global patterns of verticality because more extreme climates in the canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats in regions where macroclimate conditions are also more extreme [@oliveiraVerticalStratificationInfluences2019]. Consistently high precipitation facilitates verticality, while temperatures at both hot and cold extremes and low or highly seasonal precipitation generally limit verticality. For example, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life strategies outside moist tropical forests. However, our models indicated nuanced influences of drivers across taxa.
